Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-07-30 summary.Rmd
still working on proof of theorem 5 and additional examples

$$\max\limits_{i, j} ||(\hat{v}^{(n)}_i)^\top \hat{v}_j^{(n)}|| = O_P\bigg( \frac{(\log n)^c}{n \sqrt{\rho_n}}\bigg)$$
where $\lim\limits_{n \to 0} \rho_n = 0$ is a sparsity parameter introduced 
by \citet{307cbeb9b1be48299388437423d94bf1}. More precisely, 
$\rho_n o(\log n / \sqrt{n})$.

@@

@@

@@

**Theorem 5**. Under assumptions 1-6 as laid out by \citet{307cbeb9b1be48299388437423d94bf1},

@@

@@

@@

In this part, we will assess Algorithm 2's performance for sample sizes 
$n = 64, 128, 256, 512, 1024$. For each sample size $n$, $100$ sets of 
$\lambda^{(kl)}$'s are drawn and for each set of parameters, one adjacency 
matrix $A$ is drawn and clustered. We will not consider clustering using the 
edge probability matrix $P$ since this will always result exact recovery of the
original clusters (up to permutation).

```{r clust_err_sim, fig.cap = 'IQR clustering error using Algorithm 1 (red) compared against subspace clustering (blue) for sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size. The $y$-axis is shown on a logarithmic scale.', cache = FALSE}
clustering.df <- readr::read_csv('clustering.csv')
clustering.df %>%
  dplyr::group_by(n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_y_log10() +
  labs(y = 'clustering error', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err,
                colour = 'Algorithm 1')) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
                    colour = 'Algorithm 1')) + 
  geom_line(aes(x = n, y = med.err.ssc,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
                    colour = 'SSC')) + 
  scale_colour_brewer(palette = 'Set1')
```

@@

```{r eval = FALSE}
set.seed(123456)

n <- 64
z <- sample(c(1, 2), n, prob = c(alpha, 1 - alpha), replace = TRUE)
z <- sort(z)

n1 <- sum(z == 1)
n2 <- sum(z == 2)

lambda11 <- rbeta(n1, a1, b1)
lambda22 <- rbeta(n2, a1, b1)
lambda12 <- rbeta(n1, a2, b2)
lambda21 <- rbeta(n2, a2, b2)

X <- cbind(c(lambda11, rep(0, n2)),
           c(lambda12, rep(0, n2)),
           c(rep(0, n1), lambda21),
           c(rep(0, n1), lambda22))
Y <- cbind(c(lambda11, rep(0, n2)),
           c(rep(0, n1), lambda21),
           c(lambda12, rep(0, n2)),
           c(rep(0, n1), lambda22))
P <- X %*% t(Y)
A <- draw.graph(P)
Z <- embedding(P)
Zhat <- embedding(A, 3, 1)
```

Setting $n = 64$ and generating one example set of $\{\lambda^{(kl)}\}$, we can 
see the latent positions $XU$ follow the expected properties 
(Fig. \ref{fig:xu}).

```{r xu, fig.cap = 'Latent positions $XU$ used to construct the PABM edge probability matrix $P$ under the GRDPG framework. Note that the the points follow the properties outlined in Section 3.2.', eval = FALSE}
perm <- cbind(c(1, 0, 0, 0),
              c(0, 0, 1, 0),
              c(0, 1, 0, 0),
              c(0, 0, 0, 1))
U <- eigen(perm)$vectors[, c(3, 1, 2, 4)]
XU <- X %*% U

pairs(XU, asp = 1, col = z)
```

If we instead take the ASE of $P = X \Pi X^\top$, we fail to obtain an 
embedding that has the same properties (Fig. \ref{fig:ase_of_p}).

```{r ase_of_p, fig.cap = 'The adjacency spectral embedding of the edge probability matrix $P$. Note that even though $P = (XU) I_{p, q} (XU)^\\top$, the resulting embedding does not follow the same properties of $XU$, even up to rotation.', eval = FALSE}
pairs(Z, asp = 1, col = z)
```

Denoting $Z$ as the ASE of $P$, we can recover $ZQ$ that is equivalent to $XU$ 
(Fig. \ref{fig:q_trans}).

```{r q_trans, fig.cap = 'Pair plots of $ZQ$ where $Q$ was found by minimizing objective function (1). This embedding is equivalent to $XU$ up to 90-degree rotations and reflection along axes.', cache = TRUE, eval = FALSE}
XU.prime <- transform.embedding(Z, z, runif(4, -pi, pi))
pairs(XU.prime, col = z, asp = 1)
```

The ASE of the adjacency matrix $A$ (as one realization of $P$) has the 
resulting pairs plot:

```{r a_ase, fig.cap = 'Pair plots of the adjacency spectral embedding of $A$ Since $K=2$ is fixed, we use the signature $(3, 1)$ in this embedding.', eval = FALSE}
pairs(Zhat, col = z, asp = 1)
```

Minimizing the objective function in (1) results in the following embedding:

```{r q_from_a, cache = TRUE, fig.cap = 'Pair plots of $\\hat{Z} $ where $\\hat{Z}$ is the ASE of an adjacency matrix of a PABM and $Q$ is estimated by minimizing objective function (1).', eval = FALSE}
XU.hat <- transform.embedding(Zhat, z, runif(4, -pi, pi))
pairs(XU.hat, col = z, asp = 1)

# X.hat <- XU.hat %*% t(U)
# plot(abs(X.hat) > .25)
```

Figure \ref{fig:lambda_est} shows the medians and interquartile ranges of root 
mean square errors for Algorithm 4 over 100 simulations using the same 
parameters as before. Comparison against an MLE-based method 
\cite{noroozi2019estimation} suggests similar performance.

```{r lambda_est, cache = FALSE, fig.cap = 'Log-log plot of the median and IQR RMSE from Algorithm 4 (red) compared against an MLE-based method (blue) using sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.'}
rmse.df <- readr::read_csv('rmse.csv')

@@

>>>>

$$\max\limits_{i, j} ||(\hat{v}^{(n)}_i)^\top \hat{v}_j^{(n)}|| = O_P\bigg( \frac{(\log n)^c}{n \sqrt{\rho_n}}\bigg)$$
where $\lim\limits_{n \to 0} \rho_n = 0$ is a sparsity parameter introduced 
by \citet{307cbeb9b1be48299388437423d94bf1}. More precisely, 
$\rho_n = o(\log n / \sqrt{n})$.

@@

This leads to the following algorithm:

@@

Theorem 4 implies that as $n \to \infty$, the number (not just proportion) of 
misclassified vertices, up to permutation, outputted by algorithm 2 goes to 0.

@@

**Theorem 5**. Under assumptions 1-6 as laid out by \citet{307cbeb9b1be48299388437423d94bf1}, the estimators outputted by 
algorithm 4 converge to the true values as follows:

@@

The convergence rate of algorithm 4 is the same as that of the methods 
described by \citet{307cbeb9b1be48299388437423d94bf1}.

@@

## Case $K = 2$

@@

### Clustering

In this part, we will assess Algorithm 2's performance for sample sizes 
$n = 64, 128, 256, 512, 1024, 2048$. For each sample size $n$, $100$ sets of 
$\lambda^{(kl)}$'s are drawn and for each set of parameters, one adjacency 
matrix $A$ is drawn and clustered. We will not consider clustering using the 
edge probability matrix $P$ since this will always result exact recovery of the
original clusters (up to permutation).

```{r clust_err_sim, fig.cap = 'IQR clustering error using Algorithm 2 (red) compared against subspace clustering (blue) for sample sizes from 64 to 2048. Simulations were repeated 100 times for each sample size. The $y$-axis is shown on a logarithmic scale.', cache = FALSE}
clustering.df <- readr::read_csv('clustering.csv')
clustering.df %>%
  dplyr::group_by(n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_y_log10() +
  # scale_x_log10() + 
  labs(y = 'misclustering proportion', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
                    colour = 'Algorithm 2')) + 
  geom_line(aes(x = n, y = med.err.ssc,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
                    colour = 'SSC')) + 
  scale_colour_brewer(palette = 'Set1')
```

Theorem 4 implies that algorithm 2 will result in not just in the error rate 
converging to 0 but the error *count* as well (Fig. \ref{fig:clust_err_ct_sim}).

```{r clust_err_ct_sim, fig.cap = 'IQR of counts of misclustered vertices using Algorithm 2 (red) compared against subspace clustering (blue) for sample sizes from 64 to 2048. Simulations were repeated 100 times for each sample size.', cache = FALSE}
clustering.df <- readr::read_csv('clustering.csv')
clustering.df %>%
  dplyr::group_by(n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  # scale_y_log10() +
  # scale_x_log10() + 
  labs(y = 'misclustering count', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Algorithm 2')) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'SSC')) + 
  scale_colour_brewer(palette = 'Set1')
```

@@

### Parameter estimation

Figure \ref{fig:lambda_est} shows the medians and interquartile ranges of root 
mean square errors for Algorithm 4 over 100 simulations using the same 
parameters as before. Comparison against an MLE-based method 
\cite{noroozi2019estimation} suggests similar performance, close to the rate 
$n^{-1/2}$.

```{r lambda_est, cache = FALSE, fig.cap = 'Log-log plot of the median and IQR RMSE from Algorithm 4 (red) compared against an MLE-based method (blue) using sample sizes from 64 to 2048. Simulations were repeated 100 times for each sample size.'}
rmse.df <- readr::read_csv('rmse.csv')

@@

## Higher values of $K$

**TODO**
